---
title: "Practice Activity Week 1"
author: "Emi Degembe - STAT 541"
date: today
format: 
    html: 
        warning: false
        message: false
embed-resources: true
editor: source
code-tools: true
---

```{r}
#| label: packages-setup
#| warning: false
#| echo: false 

library(tidyverse)
```

## Quarto

First, let's make sure you know how to use Markdown formatting to style a Quarto
document.

1.  **Make this text bold.**

2.  *Make this text italicized.*

3.  Make these items into a bullet point list:

* Apples 
* Bananas 
* Potatoes

4.  Edit the YAML to remove warning messages from being output in the rendered
HTML file

5.  Using code chunk options, make it so this chunk shows the code but not the
output:

```{r}
#| label: simple-math
#| output: false

2 + 2
```

6.  Using code chunk options, make it so the file can still render even though
this chunk has an error

```{r}
#| label: code-with-error
#| eval: false

2 + a
```

## Data Visualization Review

In this class, we will exclusively make visualizations using `ggplot2` and 
packages developed to be compatible with `ggplot2`. If you are less familiar
with plotting in `ggplot2` or it's been some time since you used this package, 
I would strongly recommend referencing [resources from the data visualization section of this week's coursework](../1810-review/ggplot-review.qmd)!  

The questions below (Questions 7-11) utilize the `mpg` data from the 
`tidyverse` package.

7.  Use side-by-side boxplots to visualize the distribution of `hwy` miles per
gallon for each `manufacturer`. **The names of the manufacturers should be easily readable, and your axes should include any necessary information regarding the variables and their units!**

```{r}
#| label: side-by-side-boxplots

head(mpg)

ggplot(data = mpg, aes(x = manufacturer, y = hwy)) +
    geom_boxplot() + 
    theme_minimal() + 
    scale_x_discrete(guide = guide_axis(n.dodge = 2)) +
    labs(
        x = "Car Manufacturer", 
        y = "", 
        title = "Side-by-Side Comparison of Highway Miles per Gallon (mpg) by Manufacturer"
    ) 
```

8. Use faceted scatterplots to visualize how the relationship between `hwy`
mileage and `cty` mileage has changed over time. **Your axes should include any necessary information regarding the variables and their units!**

```{r}
#| label: faceted-scatterplots

ggplot(data = mpg, aes(x = cty, y = hwy)) +
    geom_point() + 
    facet_wrap(year ~ ., ncol = 3) + 
    labs(
        x = "City mpg", 
        y = "Highway mpg", 
        title = "Mpg Relationship Over Time", 
        subtitle = "(Mpg = miles per gallon)"
    ) + 
    theme_minimal() + 
    theme(
        axis.title.y = element_text(angle = 0)
    )
```

## Data Wrangling Review

In this class, we will exclusively make use of tools from the 
`tidyverse` suite of packages to perform our data cleaning and wrangling
operations. If you are less familiar with these packages or it's been some time
since you used them, I would strongly recommend referencing the function 
documentation!  

For these problems, we will continue to work with the `mpg` data frame, 
making various changes to the data to clean it up. 

9. The `fl` variable describes the type of fuel for each car, with levels: `p`, 
`r`, `e`, `d`, and `c`. Do some research into what each of these labels mean! 
Then, use the `mutate()` and `if_else()` functions to create a new variable
(`fuel_type`) with two levels: `petrol` (any car using petrolium-based gas) and
`alternative energy` (any car not using petrolium-based gas). 

```{r}
#| label: creating-fuel-type-variable

#fl = fuel type, 
mpg |>
    mutate(fl = if_else(fl %in% c("p", "r", "d"), "petrol", "alternative energy")) |>
    head()
```

10. The `drv` variable describes if the car has front drive (`f`), rear drive
(`r`), or four wheel drive (`4`). Let's make better labels for these values! 
Specifically, use the `mutate()` and `recode_values()` to change the `drv`
variable to have the following descriptive levels: `front`, `rear`,
`four wheel`. 

```{r}
#| label: descriptive-drive-variable

mpg |>
    mutate(drv = recode_values(drv,
        "f" ~ "front", 
        "r" ~ "rear", 
        "4" ~ "four wheel"
    )) |>
    head()

```

11. The `trans` variable contains two pieces of information, (1) the
transmission style (`auto` or `manual`) and the specific type of transmission
(e.g., `l5`, `m5`, `av`). Using the `separate_wider_delim()` function, create
two new variables from `trans`, (1) `trans_type` containing the specific type of
transmission of each car (e.g., `l5`), and (2) `trans_style` the style of the 
transmission (e.g., `auto`). *Hint: You will need to deal with the stray parenthesis!* 

```{r}
#| label: separating-transmission-variables

mpg |>
    separate_wider_delim(trans, delim = "(", names = c("trans_style", "trans_type")) |>
    mutate(trans_type = str_remove(trans_type, pattern = '\\)')) |>
    head()
    
```

## Getting to know your classmates

12. Find someone who took Stat 331 from a different professor than you. Compare
your experiences. Tell me their name and professor. List one or two things that
you think you learned more about, and one or two things that they learned more
about.

Alvaro Ramos took STAT 331 with Dr. Robinson. One thing that they learned more about was more in depth on the accessibility of data visualizations. One thing that my section had was enforced version control and creating a portfolio which we defended for a grade that we believed in. 

13. Find someone in the class who **does not** share your birth month. Tell me
their name and birthday, and use R to find out how many days apart your
birthdays are.

```{r}
#| label: birthday-calculation

a_birth <- ymd(20031230) #Dec 30, 2003
e_birth <- ymd(20050829) #Aug 29, 2005

diff_birth <- e_birth - a_birth

paste("there is a difference of", diff_birth, "days between the birth date of Alvaro and Emi.")
```

