---
title: "Practice Activity Week 2"
format: html
embed-resources: true
editor: source
---

1. Write a function called `summ_stats()` that takes a vector `x` as
input and produces the following output:

- for **numeric variables**: returns the mean, median, standard deviation, and
IQR as a data frame
- for **categorical variables**: returns the number of levels (categories) of
the variable

*Hint:* You can use `tibble()` to create the data frame.
    * Example: `tibble(a = 1:2, b = 2:3)` creates a data frame with variables
    `a` and `b`.

```{r}
#| label: sum-stats-function
#| warning: false
#| message: false

library(tidyverse)

summ_stats <- function(x) {
    
    # if x contains numeric values, return a tibble with the mean, med, sd, IQR
    if (is.numeric(x)) {
        out_df <- tibble(
            mean = mean(x), 
            median = median(x), 
            stdev = sd(x), 
            iqr = IQR(x)
        )
    # if x contains categorical, return tibble with # of levels in variable 
    } else if (is.factor(x) || is.character(x)) {
        out_df <- tibble(
            n_levels = n_distinct(x)
        )
    } else {
        stop("Inputted values are not numeric or categorical.")
    }

    #return the data frame.
    out_df
}

```

2. Confirm that your function works on each type of variable by running the 
code below. 

```{r}
summ_stats(diamonds$carat)

summ_stats(diamonds$color)
```

3. Use `map()` to apply your `summ_stats()` function to every column in the 
`diamonds` dataset. 
*Hint 1: Look up the `bind_rows()` documentation from dplyr to combine summary statistics for all quantitative variables into one data frame.*
*Hint 2: The `.id` argument will be especially helpful in adding the variable names!*


```{r}
#| label: sum-stats-map

summary_df <- map(diamonds, .f = summ_stats) |>
    bind_rows(.id = "id")

summary_df
```

4. Let's make the output more intuitive, with the variables on the columns and
the summary statistics on the rows. 

*Hint:* You will need to do a double pivot (`pivot_longer()` then
`pivot_wider()`) to achieve this result!

```{r}
#| label: double-pivot

summary_df |>
    pivot_longer(!id, 
    names_to = "stat", 
    values_to = "value") |>
    pivot_wider(names_from = "id", 
    values_from = "value")
```

## Required for STAT 541, Optional for STAT 431

5.  Now that we have the output we want, let's use our code to write a
`summ_df()` function that takes a data frame as an input and outputs a table of
summary statistics for every variable in the data frame.
*Hint: The body of your function should contain __all__ the code from Question 4.*

```{r}
#| label: df-function

summ_df <- function(df) {
    try(if(is.data.frame(df)) stop("Inputted value is not a data frame."))

    #create the summary stat tibble combining Q3 + Q4
    map(df, .f = summ_stats) |>
        bind_rows(.id = "id") |>
        pivot_longer(!id, 
            names_to = "stat", 
            values_to = "value") |>
        pivot_wider(names_from = "id", 
            values_from = "value")

}
```

6. Demonstrate that your function works using the `mpg` dataset. 

```{r}
#| label: try-df-function

summ_df(mpg)
```
