---
title: "Practice Activity Week 4"
format: html
embed-resources: true
editor: source
---

```{r}
#| message: false
#| warning: false

library(tidyverse)
```

# Roundabouts

This week, our activity will explore how roundabouts are used around the world!
You might not be as big of a fan of roundabouts as Dr. T, but these data allow 
us to investigate a fun and unique visualization
technique---[a Sankey diagram](https://en.wikipedia.org/wiki/Sankey_diagram).

A Sankey diagram visualizes data **flows**. Specifically, the visual models the
movement from one state to another. We are going to use a Sankey plot to 
visualize how traffic intersections are transformed into various types of 
roundabouts. 

## The `roundabouts` Package

The [`roundabouts` R package](https://github.com/EmilHvitfeldt/roundabouts) 
provides a simple way to interact with the 
[Roundabouts Database](https://roundabouts.kittelson.com/). You likely do not
have the package installed on your computer, so the first step is to install
the package:

```{r}
#| eval: false
#| label: install-roundabouts

pak::pak("emilhvitfeldt/roundabouts")
```

The `roundabouts` dataframe that is included in the package has an `address`
variable which is a bit untidy. Below is some data cleaning code that will 
clean the `address` variable, and create three new variables: `town_city`, 
`county_area`, and `state_region`. 

```{r}
#| label: roundabouts-data-cleaning
#| message: false
#| warning: false

library(roundabouts)

roundabouts_clean <- roundabouts |>
  mutate(
    # Step 1: Remove strange <![CDATA[ tags from the beginning of the address field
    address = stringr::str_remove(address, stringr::fixed("<![CDATA[")), 
    # Step 2: Remove strang ]]> tags from the end of the address field
    address = stringr::str_remove(address, stringr::fixed("]]>")),
    # Step 3: Isolate the country at the end of the string between the ( ) symbols
    country = stringr::str_extract(address, "\\([^)]+\\)$"),
    # Step 4: Remove the brackets surrounding the country name
    country = stringr::str_remove_all(country, "[\\(\\)]"), 
    # Step 5: Remove the country from the address field
    address2 = stringr::str_remove(address, "\\s*\\([^)]+\\)$")) |>
  # Step 6: Separate address into town, county, state by comma
  separate(address2, 
                  into = c("town_city", "county_area", "state_region"), 
                  sep = ",") |> 
  # Step 7: Fix one instance of California instead of CA
  mutate(state_region = str_trim(state_region, side = "both"), 
         state_region = fct_recode(state_region, 
                                   "CA" = "California")
         ) |> 
  # Step 8: Remove companies from states
  filter(! str_detect(state_region, pattern = "Co.")) |> 
  # Step 9: Grab columns we will be using
  select(town_city, county_area, state_region, country, 
         year_completed, type, previous_control_type, control_type)

```

## Creating a Region Variable

Alright, we now have the `country` of each roundabout, but that seems to 
fine grain for our analysis. If we're interested in how roundabouts are used
around the world, it seems like a `region` variable might be a better fit. 

1. Create a new `continent` column in the dataset, which classifies each 
`country` into one of the five major continents---Americas, Europe, Asia, 
Oceania, Africa.

The `countrycode()` function from the `countrycode` package will be helpful
here! Here is an example of how you can use this function to obtain the 
continent of a given country:

```{r}
#| label: continent-demo
#| eval: false

countrycode::countrycode("Brazil",
            origin = "country.name",
            destination = "continent")
```

```{r}
#| label: add-continent-column

roundabouts_clean <- roundabouts_clean |>
  mutate(continent = countrycode::countrycode(country,
                                              origin = "country.name",
                                              destination = "continent"))

```

## Exploring Roundabouts

Let's explore the data a bit before we jump into visualizing these data. 

2. What countries have the most roundabouts (in these data)? The fewest
roundabouts?

```{r}
#| label: country-total-roundabouts

# determining the most roundabouts

roundabouts_clean |>
  group_by(country) |>
  summarise(count = n()) |>
  arrange(desc(count)) |>
  head(5)

# determining the least roundabouts

roundabouts_clean |>
  group_by(country) |>
  summarise(count = n()) |>
  arrange(count) |>
  head(5)

```

3. In these data, what types of roundabouts are most common?

```{r}
#| label: common-types

roundabouts_clean |>
  group_by(type) |>
  summarise(count = n()) |>
  arrange(count) |>
  head(3)

```

4. How does the distribution of types differ by continent?
*Hint: Proportions are the easiest way to address this question!*

```{r}
#| label: types-by-continent

# rows = the type cols = each continent

roundabouts_clean |>
  select(continent, type) |>
  group_by(continent, type) |>
  summarise(count = n()) |>
  mutate(total = sum(count), 
         proportion = count/total) |>
  ungroup() |>
  select(type, continent, proportion) |>
  pivot_wider(names_from = continent, 
  values_from = proportion, 
  values_fill = 0)

```

5. We've seen a lot of traffic calming circles showing up around SLO. What
states in the US have the largest proportion of traffic calming circles?

```{r}
#| label: prop-calming-circle-in-US

roundabouts_clean |>
  filter(country == "United States" & type == "Traffic Calming Circle") |>
  group_by(state_region) |>
  summarise(count = n()) |>
  arrange(desc(count)) |>
  head(5)

```

# The `ggalluvial` Package

Now that we have the data ready, let's explore making Sankey plots with the 
[`ggalluvial` R package](https://corybrunson.github.io/ggalluvial/). 

We are interested in visualizing how roundabouts are incorporated into traffic
intersections. To do this, we will visualize the *flow* from the previous
intersection type to the type of roundabout that was added, to the final 
intersection (control) type. 

Let's start by creating the data `gglluvial` requires for plotting. 

6. Create a table of frequencies of previous intersection types, current
intersection types, and roundabouts. Let's focus on intersections (both old and
new) where the type was known (not labeled as `"Unknown"`).

```{r}
#| label: ggalluvial-summaries

roundy_sankey <- roundabouts_clean |>
  filter(type != "Unknown" & previous_control_type != "Unknown" & "control_type" != "Unknown") |>
  count(previous_control_type, control_type, type)

```

## Customizing Labels -- Required for 541, Optional for 431

7. Now create the Sankey plot using the data you made above! Your axes should 
be previous control type and the current control type, and the colors should be
associated with the type of roundabout. *Hint: You might find the [ggalluvial reference page helpful](https://corybrunson.github.io/ggalluvial/)!*

```{r}
#| label: Sankey-plot

library(ggalluvial)

roundy_sankey |>
ggplot(aes(axis1 = previous_control_type, axis2 = control_type, y = n)) +
  scale_x_discrete(
    limits = c("previous_control_type", "control_type"), 
    expand = c(.2, .05), 
    labels = c(
      "previous_control_type" = "Previous", 
      "control_type" = "Current"
    )
  ) + 
  geom_alluvium(aes(fill = type)) +
  geom_stratum() +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  labs(
    title = "Sankey Diagram of Previous to Current Roundabouts", 
    subtitle = "By counts", 
    y = ""
  ) +
  theme_minimal()

```

8. Now that you have a working solution, let's make the strata labels easy to
read. Look over this [reference article](https://corybrunson.github.io/ggalluvial/articles/labels.html) on different ways to label the strata for each axis. Implement your favorite 
method! 

```{r}
#| label: better-strata-labels

roundy_sankey |>
ggplot(aes(axis1 = previous_control_type, axis2 = control_type, y = n)) +
  scale_x_discrete(
    limits = c("previous_control_type", "control_type"), 
    expand = c(.2, .05), 
    labels = c(
      "previous_control_type" = "Previous", 
      "control_type" = "Current"
      )
    ) + 
  geom_alluvium(aes(fill = type)) +
  geom_stratum(alpha = 0.5) +
  ggrepel::geom_text_repel(
    aes(label = ifelse(after_stat(x) == 1, as.character(after_stat(stratum)), NA)),
    stat = "stratum",
    direction = "y",
    nudge_x = -1
  ) +
  ggrepel::geom_text_repel(
    aes(label = ifelse(after_stat(x) == 2, as.character(after_stat(stratum)), NA)),
    stat = "stratum",
    direction = "y",
    nudge_x = 1
  ) +
  labs(
    title = "Sankey Diagram of Previous to Current Roundabouts", 
    subtitle = "By counts", 
    y = ""
  ) +
  theme_minimal()

```

## Customizing Colors -- Required for Both 541 and 431

9. Final touches! Let's add some custom colors and an informative plot title!
*Each of the axes in your plot should have a title as well!*

```{r}
#| label: final-touches
library(ggfittext)

pal <- c("#D55E00", "#E69F00", "#F0E442", "#019E73", "#0072B2")

roundy_sankey |>
  ggplot(aes(axis1 = previous_control_type, axis2 = control_type, y = n)) +
  scale_x_discrete(
    limits = c("previous_control_type", "control_type"),
    expand = c(.2, .05),
    labels = c(
      "previous_control_type" = "Previous",
      "control_type" = "Current"
    )
  ) +
  geom_alluvium(aes(fill = type)) +
  geom_stratum(alpha = 0.5) +
  scale_fill_manual(values = pal) +
  ggrepel::geom_text_repel(
    aes(label = ifelse(after_stat(x) == 1, as.character(after_stat(stratum)), NA)),
    stat = "stratum",
    direction = "y",
    nudge_x = -1
  ) +
  ggrepel::geom_text_repel(
    aes(label = ifelse(after_stat(x) == 2, as.character(after_stat(stratum)), NA)),
    stat = "stratum",
    direction = "y",
    nudge_x = 1
  ) +
  labs(
    title = "Sankey Diagram of Previous to Current Control Method Based on Roundabout Type",
    subtitle = "By counts",
    y = ""
  ) +
  theme_minimal()

```
